---
title: "Getting Started with bibnets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with bibnets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bibnets)
```

## Introduction

`bibnets` imports scholarly metadata and constructs bibliometric networks
from works, authors, references, keywords, sources, countries, and
institutions. It represents co-authorship, co-citation, bibliographic
coupling, keyword co-occurrence, direct citation, historiograph, and custom
co-occurrence analyses as network edge lists.

### Input and imports

`read_biblio()` is the entry point for file-based input. It accepts a file, a
set of files, or a directory, detects supported file exports when
`format = "auto"`, and dispatches to the corresponding reader. The file
readers cover Scopus CSV with `read_scopus()`, Web of Science plain-text and
tab-delimited exports with `read_wos()`, flat OpenAlex CSV with
`read_openalex_csv()`, BibTeX with `read_bibtex()`, RIS with `read_ris()`,
Lens.org CSV with `read_lens()`, Dimensions CSV with `read_dimensions()`,
and generic CSV input through `read_biblio(format = "generic")` or inferred
generic input when source columns are mapped by name. OpenAlex records
returned by `openalexR::oa_fetch()` are converted with `read_openalex()`.
Crossref records returned by `rcrossref::cr_works()` are converted with
`read_crossref()`.

The source-specific readers return a common data frame with `id`, `title`,
`year`, `journal`, `doi`, `cited_by_count`, `abstract`, `type`, and
list-columns such as `authors`, `references`, and `keywords`.
Source-specific fields, including `countries`, `affiliations`,
`index_keywords`, and `keywords_plus`, are kept when present.

### Network construction

Network construction follows the same matrix path across builders. A
works-by-entities bipartite matrix is created by `build_bipartite()` or, for
position-dependent weighting, by `build_author_bipartite()`. The matrix is
weighted by `apply_counting()` or by positional weights, projected by
`multiply_bipartite()`, and returned by `as_bibnets_network()`.

The counting methods are `"full"`, `"fractional"`, `"paper"`, `"strength"`,
`"harmonic"`, `"arithmetic"`, `"geometric"`, `"adaptive_geometric"`,
`"golden"`, `"first"`, `"last"`, `"first_last"`, and
`"position_weighted"`. The attention profiles are `"lead"`, `"last"`,
`"proximity"`, and `"circular"`; in the implementation these map to
`"attention_lead"`, `"attention_last"`, `"attention_proximity"`, and
`"attention_circular"` positional weights. Attention weighting is available
for author, keyword, country, and institution networks.

Projected matrices can be returned as raw counts with `"none"` or normalized
with `"cosine"`, `"association"`, `"jaccard"`, `"inclusion"`, or
`"equivalence"`. The same similarity names are accepted by `normalize()`.

### Speed

The incidence matrix is stored as a sparse `dgCMatrix`. Projection uses
`crossprod()` for column-mode networks and `tcrossprod()` for row-mode
networks, and edge extraction is performed from the sparse projected matrix
without materializing a dense matrix. Memory use follows the number of
non-zero matrix entries and retained edges rather than the size of a dense
node-by-node matrix. The hard dependencies are Matrix, stats, and utils.

### Exports

The default return value is an edge list. Export and conversion functions are
`to_igraph()`, `to_tbl_graph()`, `to_cograph()`, `to_gephi()`,
`to_graphml()`, and `to_matrix()`. `to_gephi()` writes or returns
Gephi-ready node and edge CSV tables. `to_graphml()` writes or returns
GraphML. `to_matrix()` returns a sparse adjacency matrix.

### Output

Every builder returns a **`bibnets_network`**: a tidy data frame with four
columns —

- `from`, `to` — the two endpoints of an edge,
- `count` — the raw binary co-occurrence count for that pair,
- `weight` — the analytical weight after counting and optional similarity
  normalization.

With `counting = "full"` and `similarity = "none"`, `weight` equals
`count`. They diverge once fractional counting or a similarity measure is
applied.

The builders, at a glance:

| Function | Nodes | An edge means |
|---|---|---|
| `author_network()` | authors | co-authorship, author coupling, or co-citation |
| `reference_network()` | cited references | two references cited together |
| `document_network()` | documents | shared references, shared citers, or direct citation |
| `keyword_network()` | keywords | two keywords appear together |
| `source_network()` | journals | sources share references or are co-cited |
| `country_network()` | countries | countries collaborate or share references |
| `institution_network()` | institutions | institutions collaborate or share references |
| `conetwork()` | any field | entities co-occur, or share values of another field |
| `local_citations()` | documents | within-corpus citation counts |
| `historiograph()` | documents | directed citation history among top-cited papers |
| `temporal_network()` | any builder's nodes | the same network over time windows |

## Quick start

You do **not** need a special reader. Any data frame with one row per paper
works — point a builder at the column that holds the entity and tell it the
delimiter:

```{r quick-df}
papers <- data.frame(
  `Author Names` = c("Smith J, Doe A, Lee K", "Smith J, Lee K",
                     "Doe A, Lee K", "Smith J, Doe A"),
  check.names = FALSE
)

author_network(papers, authors = "Author Names", sep = ",")
```

If your data is a scholarly export instead, read it first — the format is
detected from the file content — then build with the defaults:

```{r quick-reader, eval = FALSE}
data    <- read_biblio("scopus.csv")
authors <- author_network(data, type = "collaboration")
```

Either way the result is the same four-column edge list, ready to inspect,
prune, or export.

## Reading your own data

### Scholarly exports

`read_biblio()` accepts a file, a folder, or several files, and detects
Scopus, Web of Science, OpenAlex, BibTeX, RIS, Lens.org, and Dimensions
from the content:

```{r read-files, eval = FALSE}
data <- read_biblio("export.csv")
data <- read_biblio("folder_with_exports/")
data <- read_biblio(c("part_1.csv", "part_2.csv"))
```

The format-specific readers can also be called directly
(`read_scopus()`, `read_wos()`, `read_openalex_csv()`, `read_dimensions()`,
`read_lens()`, `read_bibtex()`, `read_ris()`).

### A custom CSV

For a CSV that matches no known export, map each source column onto a
standard field **by name** — `authors`, `keywords`, `references`,
`countries`, `affiliations`, or `journal`. Naming any of them reads the
file as a generic CSV, so you do not pass `format` yourself:

```{r read-generic, eval = FALSE}
data <- read_biblio(
  "custom.csv",
  id       = "paper_id",
  authors  = "Author Names",
  keywords = "Tags",
  sep      = ","
)
```

Each mapped column is split on `sep` into the standard list-column, so
afterwards every builder works with its defaults.

### A plain data frame, directly

As the quick start showed, you can skip the reader entirely and let the
builder split a column for you. The same column arguments are available on
every builder:

```{r read-direct, eval = FALSE}
author_network(my_df, authors = "Author Names", sep = ",")
keyword_network(my_df, keywords = "Tags",       sep = ",")
```

The work identifier is the `id` column. You need not supply one: when no
`id` column is present each row is treated as one document; pass
`id = "paper_id"` to use a differently-named column. Surrounding quotes are
stripped by default (`strip_quotes = TRUE`), and in a coupling network the
references column takes its own `references_sep`. The companion
`vignette("reading-data")` covers every reader and these options in full.

### The standard schema

Readers return a common set of columns:

```{r schema}
data(scopus_quantum_cloud)
sc <- scopus_quantum_cloud
names(sc)[1:12]
```

The columns that matter for network construction are `id`, the list-columns
`authors` / `references` / `keywords`, and `year` (used by
`temporal_network()`). Source-specific extras such as `countries`,
`affiliations`, and `keywords_plus` are kept when available.

## Datasets used here

```{r data}
data(biblio_data)
data(learning_analytics)

small <- biblio_data            # tiny, synthetic
oa    <- learning_analytics     # 1,508 OpenAlex records on learning analytics

c(small = nrow(small), scopus = nrow(sc), openalex = nrow(oa))
```

## Author collaboration

Two authors are linked when they appear on the same paper:

```{r author-basic}
authors <- author_network(oa, type = "collaboration")
head(authors, 5)
summary(authors)
```

Use `min_occur` to drop rare authors before projection:

```{r author-minoccur}
nrow(author_network(oa, type = "collaboration"))
nrow(author_network(oa, type = "collaboration", min_occur = 2))
```

### Counting methods

`counting` controls how much each paper contributes to an edge:

```{r counting}
head(author_network(small, type = "collaboration", counting = "full"), 3)
head(author_network(small, type = "collaboration", counting = "fractional"), 3)
head(author_network(small, type = "collaboration", counting = "harmonic"), 3)
head(author_network(small, type = "collaboration", counting = "first_last"), 3)
```

- `"full"` — every observed co-occurrence counts 1 (use for raw events).
- `"fractional"` — scaled by list size so large teams don't dominate.
- `"harmonic"` / `"first_last"` and other position methods — only when
  author order is analytically meaningful.

### Attention weights

`attention` applies a smooth positional profile instead of a named counting
scheme (available for author, keyword, country, and institution networks):

```{r attention}
head(author_network(small, attention = "lead"), 3)
```

| `attention` | Highest weight at |
|---|---|
| `"lead"` | first position |
| `"last"` | last position |
| `"proximity"` | middle positions |
| `"circular"` | first and last |

## Reference co-citation

Two references are linked when a paper cites both:

```{r cocitation}
refs <- reference_network(sc, min_occur = 2)
head(refs, 5)
```

A similarity measure offsets the advantage of very frequently cited works:

```{r cocitation-cosine}
head(reference_network(sc, min_occur = 2, similarity = "cosine"), 3)
```

## Document coupling and citation

Coupling links two documents that share cited references:

```{r coupling}
head(document_network(sc, type = "coupling", similarity = "cosine"), 5)
```

Direct citation is directed — `from` cites `to` — and only within the
corpus (the cited work must also be a row in the data):

```{r citation}
head(document_network(sc, type = "citation"), 5)
```

## Keyword co-occurrence

```{r keywords}
kw <- keyword_network(sc, min_occur = 2)
head(kw, 5)
```

Labels are trimmed and upper-cased during construction, so
`machine learning`, `Machine Learning`, and ` MACHINE LEARNING ` are one
node. Association strength is a common choice for co-occurrence maps:

```{r keywords-assoc}
head(keyword_network(sc, min_occur = 2, similarity = "association"), 3)
```

## Countries, institutions, and sources

```{r geo}
head(country_network(oa, counting = "fractional"), 5)
head(institution_network(oa, counting = "fractional", min_occur = 2), 5)
head(source_network(sc, type = "coupling", min_occur = 2), 5)
```

For coupling networks, `min_occur` is applied to the aggregated entity
before the network is built.

## Generic co-networks

`conetwork()` covers projections without a dedicated wrapper. One field
links entities that co-occur; a second field (`by`) links them through a
shared value:

```{r conetwork}
head(conetwork(sc, "keywords", min_occur = 2), 3)
head(conetwork(sc, "authors", by = "keywords", min_occur = 2), 3)
```

The second result links authors through shared keywords — a thematic
similarity network, not a co-authorship one.

## Normalization

The same raw counts support different similarity scores; only `weight`
changes, `count` does not:

```{r normalize}
none <- keyword_network(sc, min_occur = 2, similarity = "none")
cos  <- keyword_network(sc, min_occur = 2, similarity = "cosine")
head(none[, c("from", "to", "weight", "count")], 3)
head(cos[,  c("from", "to", "weight", "count")], 3)
```

Use `"none"` for raw weights, `"cosine"` for overlap scaled by size,
`"association"` to emphasize pairs that co-occur more than expected, and
`"jaccard"` / `"inclusion"` / `"equivalence"` to match a downstream
convention.

## Reducing large networks

```{r reduce}
edges <- author_network(oa, type = "collaboration")
c(all        = nrow(edges),
  threshold  = nrow(prune(edges, threshold = 2)),
  top_n      = nrow(prune(edges, top_n = 5)),
  top_nodes  = nrow(filter_top(edges, n = 50)))
```

- `prune(threshold = x)` — absolute edge-weight cutoff.
- `prune(top_n = k)` — keep each node's strongest edges.
- `filter_top(n = k)` — keep edges among the most-connected nodes.

`backbone()` applies the disparity filter, which keeps edges that are strong
relative to a node's local strength distribution — not a global cutoff:

```{r backbone}
bb <- backbone(edges, alpha = 0.05)
nrow(bb)
```

## Temporal networks

`temporal_network()` runs any builder over time windows (fixed, sliding, or
cumulative):

```{r temporal}
tn <- temporal_network(oa, author_network, "collaboration", window = 3)
names(tn)
```

Each window's edge list carries a `window` column. Windows with fewer than
two records, or no surviving edges, are dropped; a builder error inside a
window becomes a warning labelled with that window.

## Local citations and historiographs

`local_citations()` counts how often each document is cited by others in the
same corpus; `historiograph()` builds the directed citation graph among the
top-cited documents:

```{r historiograph}
head(local_citations(sc), 5)

h <- historiograph(sc, n = 10)
h$nodes
head(h$edges, 5)
```

Both require reference strings or IDs to match document IDs in the data; if
the cited works are external, local counts stay low.

## Author-name normalization

`parse_names()` reorders and splits author names (it recognizes
`Last, First`, `SURNAME Initials`, and `First Last`). Because node identity
is fixed when a network is built, normalize *before* building so that two
spellings of one author merge:

```{r parse-names}
parse_names(c("Saqr, Mohammed", "WANG Y", "Mohammed Saqr"))
```

See `vignette("parsing-author-names")` for the full treatment.

## Exporting

The edge list is already usable; converters cover the common targets:

```{r export}
edges <- keyword_network(sc, min_occur = 2)

m <- to_matrix(edges)            # sparse adjacency matrix
m[1:4, 1:4]

gephi <- to_gephi(edges)         # Gephi node/edge tables
head(gephi$edges, 3)

cat(substr(to_graphml(edges), 1, 200))   # GraphML, no XML dependency
```

`to_igraph()`, `to_tbl_graph()`, and `to_cograph()` are available when their
(suggested) packages are installed.

## Reading a `bibnets_network`

The object records how it was built, as attributes:

```{r attrs}
edges <- author_network(oa, type = "collaboration", counting = "harmonic")
c(type     = attr(edges, "network_type"),
  counting = attr(edges, "counting"),
  sim      = attr(edges, "similarity"))

summary(edges)
```

`print()` reports the network type, node and edge counts, and the counting
and similarity methods — so a saved edge list always says how it was made.
